Honors Thesis Proposal

### 2009-03-19 honors_thesis_proposal/paper.tex
Some updates to my thesis proposal

I will study the feasibility and potential performance gain of speculative
execution in particle swarm optimization.  By speculative execution I mean the
use of additional computational resources, presumably extra processors running
in parallel, to perform two or more iterations of particle swarm optimization
at a time, speeding up the algorithm significantly.

@@

In running experiments, many trials will be run and averaged, as there is a
random component in this algorithm.  I will use statistical tests for all
comparisons to the original PSO algorithm.  The success criterion will be a
better performance of the algorithm as measured by the lowest function
evaluation seen over a number of function evaluations for several benchmark and
real-world problems.  Speculative execution takes more computation than parallel
PSO with the same number of particles, so the two algorithms will be compared
at points where they use equal amounts of function evaluations.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart.  It tries to intelligently search a multi-dimensional space by
mimicking the swarming and flocking behavior of birds and other
animals.~\cite{kennedy-icnn95} It is a sociological algorithm that depends on
interaction between particles to quickly and consistently find the optimal
solution to a problem.  The algorithm keeps track of a number of potential
solutions, called particles, which move somewhat randomly through the search
space.  Particles remember the best place they have been, or solution they have
evaluated, and are attracted back to that place, as well as to the best
solution other particles have seen.  Specifically, the formulas for updating
the position $\ppos_t$ and velocity $\pvel_t$ of a particle at iteration $t$ are
as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\gbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ is the best position the current particle has seen,
and $\gbest$ is the best position any of the other particles have seen.  \(
\coeff_1 \), \( \coeff_2 \), and \( \constriction \) are parameters with
prescribed values required to ensure convergence (2.05, 2.05, and .73,
respectively).~\cite{clerc-tec02}~\cite{poli-aea08}

@@

In a lot of optimization algorithms, speculative execution is impossible,
because the next iteration of the algorithm depends on the evaluation of the
function during the previous iteration.  In simulated annealing, for instance,
the probability of accepting a new position depends on the difference in
function value between the new position and the current one.  In gradient
descent, the gradient of the function must be determined (which includes
actually evaluating it) to know how far to move before sampling the function
again.  But particle swarm optimization is rather unique in that it doesn't
require the evaluation of the function to know what the next position of each
particle will be.  It simply depends on \emph{which} particle ended up having
the best position and whether or not the current particle found a better spot
than it had seen before.  That means you can just compute all possible next
positions---all combinations of which particle is the new gbest and whether or
not pbest was updated (\gbest and \pbest from \eqref{eq:velupdate})---and
evaluate each position in parallel.  That computes two iterations at the same
time.  It uses a lot of extra work to get the second iteration, but for some
functions more iterations are worth more than extra particles.

@@

\begin{tabular}{|l|l|}
\hline
Run preliminary tests to ensure feasibility & February 28, 2009 \\
\hline
Finish writing and debugging code & March 30, 2009 \\
\hline
Finish gathering data and statistics & June 30, 2009 \\
\hline
Complete analysis of data and begin writing thesis & September 1, 2009 \\
\hline
Submit completed copy of thesis to advisor & November 15, 2009 \\
\hline
Submit paper detailing results to CEC 2010 & December 20, 2009 \\
\hline
Final thesis and portfolio submitted & February 1, 2010 \\
\hline
Thesis defense completed & March 1, 2010 \\
\hline
Final four thesis copies submitted & March 8, 2010 \\
\hline
\end{tabular}

@@

This project will not require any outside funding.  The only expenditure
will be my own time researching, conducting experiments, and writing.

>>>>

I will study the feasibility and potential performance gain of a technique
which we call speculative execution in particle swarm optimization.  By
speculative execution I mean the use of additional computational resources,
presumably extra processors running in parallel, to perform two or more
iterations of particle swarm optimization at a time, speeding up the algorithm
significantly.

@@

In running experiments, many trials will be run and averaged, as there is a
random component in this algorithm.  I will use statistical tests for all
comparisons to the original PSO algorithm.  The success criterion will be a
better performance of the algorithm as measured by the lowest function
evaluation seen over a number of function evaluations for several benchmark and
real-world problems.  Speculative execution requires more function evaluations 
per particle than a na\"ive parallel PSO, %TODO
PSO with the same number of particles, so the two algorithms will be compared
at points where they use equal amounts of function evaluations per iteration.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart.  It tries to intelligently search a multi-dimensional space by
mimicking the swarming and flocking behavior of birds and other
animals.~\cite{kennedy-icnn95} It is a sociological algorithm that depends on
interaction between particles to quickly and consistently find the optimal
solution to a problem.  The algorithm keeps track of a number of potential
solutions, called particles, which move somewhat randomly through the search
space.  Particles remember the best place they have been, or solution they have
evaluated, and are attracted back to that place, as well as to the best
solution other particles have seen.  Specifically, the formulas for updating
the position $\ppos_t$ and velocity $\pvel_t$ of a particle at iteration $t$ are
as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\gbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called pbest) is the best position the current
particle has seen, and $\gbest$ (called gbest) is the best position any of the
other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73,
respectively).~\cite{clerc-tec02}~\cite{poli-aea08}

@@

In many optimization algorithms speculative execution is impossible, because
the next iteration of the algorithm depends on the evaluation of the function
during the previous iteration.  In simulated annealing, for instance, the
probability of accepting a new position depends on the difference in function
value between the new position and the current one.  In gradient descent, the
gradient of the function must be determined (which includes actually evaluating
it) to know how far to move before sampling the function again.  But particle
swarm optimization is rather unique in that it does not require the evaluation
of the function to know what the next position of each particle will be.  It
simply depends on \emph{which} particle ended up having the best position and
whether or not the current particle found a better spot than it had seen
before.  That means you can just compute all possible next positions---all
combinations of which particle is the new gbest and whether or not pbest was
updated (\gbest\ and \pbest\ from \eqref{eq:velupdate})---and evaluate each
position in parallel.  That computes two iterations at the same time.  It uses
a lot of extra work to get the second iteration, but for some functions
performing more iterations makes more progress than adding extra particles.

@@

\begin{tabular}{|l|l|}
\hline
Run preliminary tests to ensure feasibility & February 28, 2009 \\
\hline
Finish writing and debugging code & April 30, 2009 \\
\hline
Finish gathering data and statistics & June 30, 2009 \\
\hline
Complete analysis of data and begin writing thesis & September 1, 2009 \\
\hline
Submit completed copy of thesis to advisor & November 15, 2009 \\
\hline
Submit paper detailing results to CEC 2010 & December 20, 2009 \\
\hline
Final thesis and portfolio submitted & February 1, 2010 \\
\hline
Thesis defense completed & March 1, 2010 \\
\hline
Final four thesis copies submitted & March 8, 2010 \\
\hline
\end{tabular}

@@

This project will not require any outside funding.  The only expenditure will
be my own time researching, conducting experiments, and writing.  I have
received a grant from the Office of Research and Creative Activities to perform
this research, for which I am grateful.
